---
title: "Code from paper 'Operative versus non-operative management of rib fractures in flail chest after cardiopulmonary resuscitation', Dorn et al."
author: "Simone Oberhaensli, Interfaculty Bioinformatics Unit, University of Bern"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    toc: true
    toc_depth: 3
  fontsize: 11pt

---




## Introduction
For all analyses a significance threshold of p < 0.05 is applied.

```{r preable, eval=F, echo=F}
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
```




```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# load libraries
library("ggplot2")
library("tidyr")
library("RColorBrewer")
library("pca3d")
library("ggpubr")
library("lmerTest")
library("car")
library("lmtest")


# settings for boxplots with bullets and labels
xnudge1 = 1.4
xnudge2 = 0.1
sz1 = 3
sz2 = 1

```



```{r data_prep, eval=T, echo=F}

# Read data from RDS object
# These are the original data, except that palliative care patients are already 
# removed
dat <- readRDS("/Users/simone/Projects/misc_short_projects/Dorn_Data_osteosynthesis/ANOVA/p555_data.rds") # Todo: Adjust this part later!!!


# for better overview create dataframe with most relevant variables for the analysis
dat.t <- dat[,c("pairs", "group","age", "frac.Sternum", "Lung.contusion", "GCS", 
                 "Respirator.time", "Resp.Time.after.OR", "Tracheotomy", 
                 "Pneumonia", "Neurology", "GI.bleeding", "Death", "female")]

# the lines below are needed to analyse time after OP for Sx group
# for this we first need to create a new variable which consists of the total 
# time for noSX group and the time after surgery for the Sx group
dat.t$newtime <- as.vector(cbind(c(dat.t[1:10,"Respirator.time"], 
                                   dat.t[11:23,"Resp.Time.after.OR"])))

# some statistics regarding cohort size
n.tot.obs <- nrow(dat.t)
n.nodec.obs <- nrow(dat.t[dat.t$Death != "1",])

```


## Analyses with both groups as one cohort

In the following section we perform some comparisons between the group that underwent 
stabilization surgery (Sx) and the group that was treated conservatively (noSx).



### Primary endpoint: time spent on respirator

Is there a significant difference in the number of days that patients spent on 
the respirator between the groups Sx and noSx?  
The cohort includes `r n.nodec.obs` individuals (deceased patients excluded).

```{r time_ttest, eval=T, echo=F, fig.height = 5, fig.width = 8}

dat.ttest <- dat.t[dat.t$Death != "1",]
cat("Data are approximatelly normally distributed")
qqPlot(dat.ttest$Respirator.time, ylab = "time on respirator", main = "Total 
       time for Sx and noSx group")
qqPlot(dat.ttest$newtime, ylab = "time on respirator", main = "noSx total time 
       and Sx time after surgery")

# for save data handling create individual dfs for each group
d.Sx <- dat.ttest[dat.ttest$group == 'Sx',]
d.nSx <- dat.ttest[dat.ttest$group == 'noSx',]

cat("There is no significant difference of total time spent on respirator between 
    the Sx and noSx group")
t.test(d.nSx$Respirator.time, d.Sx$Respirator.time)


p.tt <- ggboxplot(dat.ttest, x = "group", y = "Respirator.time",
          color = "group", palette = c("#00AFBB", "#E7B800"),
          outlier.shape = NA) + 
  geom_jitter(height = 0, width = 0.05) +
  labs(title = "Days on respirator", subtitle = "noSx total time vs Sx total time",
       x = "group",
       y = "days")
# Add pvalue
pp.tt <- p.tt + stat_compare_means(method = "t.test", 
                               paired = FALSE,
                               hjust = -0.6)



cat("There is a significant difference if the time on the respirator after 
    surgery for the Sx group is compared to the total time of the noSx group")
t.test(d.nSx$newtime, d.Sx$newtime)


p.nt <- ggboxplot(dat.ttest, x = "group", y = "newtime",
          color = "group", palette = c("#00AFBB", "#E7B800"),
          outlier.shape = NA) + 
  geom_jitter(height = 0, width = 0.05) +
  labs(title = "Days on respirator", subtitle = "noSx total time vs Sx time after surgery",
       x = "group",
       y = "days")
# Add pvalue
pp.nt <- p.nt + stat_compare_means(method = "t.test", 
                               paired = FALSE,
                               hjust = -0.6)

ggarrange(pp.tt, pp.nt, ncol = 2, nrow = 1)
```


### Secondary endpoints: Overrepresentation test of tracheotomies, pneumonia, neurological deterioration and death
The cohort includes `r n.tot.obs` individuals.  
All tested variables show now dependence from the group (Sx or noSx).

```{r fisher_tests, eval=T, echo=F}

cat("Tracheotomies are not significantly overrepresented in any of the two groups")
(t.trach <- xtabs(~ group + Tracheotomy, data = dat.t))
fisher.test(t.trach)


cat("Pneumonia is not significantly overrepresented in any of the two groups")
(t.pneu <- xtabs(~ group + Pneumonia, data = dat.t))
fisher.test(t.pneu)


cat("Neurological complications are not significantly overrepresented in any 
    of the two groups?")
(t.neu <- xtabs(~ group + Neurology, data = dat.t))
fisher.test(t.neu)


cat("Deceased patients are not significantly overrepresented in any of the two groups")
(t.death <- xtabs(~ group + Death, data = dat.t))
fisher.test(t.death)


```



## Linear regression analysis of both group as one cohort

Approach: use linear models combined with backward variable selection and likelihood ratio tests to reduce model complexity and identify predictors that have an influence on the response variable 'Respirator.time' (= total time on respirator starting from admission to ICU for both groups) and 'newtime' (time on respirator after the surgery for the Sx group), respectively.

The cohort includes `r n.nodec.obs` individuals (deceased patients excluded).

#### Linear model, data from both groups, y = total respiratory time

```{r lm_tot_time, eval=T, echo=F}
# the full model includes all predictors except comorbidities as these are not informative
# we also exclude frac.Rib.unilateral and GI bleeding because the frequency is very low
mt.full <- lm(Respirator.time ~ group + age + GCS + female + Tracheotomy + 
                frac.Sternum + Lung.contusion + 
                Pneumonia + Neurology, data = dat.t[dat.t$Death != "1",])

summary(mt.full)

# run backward selection
mt.bkw <- step(mt.full)
summary(mt.bkw)
# the reduced model is equivalent to the full model and can replace the latter
lrtest(mt.full, mt.bkw)
confint(mt.bkw)

cat("Residual plots of the reduced model")
par(mfrow = c(2,2))
plot(mt.bkw)
par(mfrow=c(1,2))

dat.nodeath <- dat.t[dat.t$Death != "1",]
g1 <- ggplot(dat.nodeath, aes(x=Tracheotomy, y=Respirator.time, label = pairs)) +
  geom_boxplot() + geom_point(pch=21, size = sz2, aes(fill=group)) + geom_text(nudge_x = xnudge2, size = sz1) +
  labs(x = "Tracheotomy no/yes", y = "days on respirator")
g2 <- ggplot(dat.nodeath, aes(x=Lung.contusion, y=Respirator.time, label = pairs)) +
  geom_boxplot() + geom_point(pch=21, size = sz2, aes(fill=group)) + geom_text(nudge_x = xnudge2, size = sz1) +
  labs(x = "Pulmonary contusion no/yes", y = "days on respirator")


```


#### Linear model, data from both groups, y = time after surgery for the Sx group and total respiratory time for noSx group


```{r lm_newtime, eval=T, echo=F, fig.height = 4, fig.width = 8}

mnt.full <- lm(newtime ~ group + age + GCS + female + Tracheotomy + 
                 frac.Sternum + Lung.contusion + Pneumonia + 
                 Neurology, data = dat.t[dat.t$Death != "1",])

summary(mnt.full)
# run backward selection
mnt.bkw <-  step(mnt.full)
summary(mnt.bkw)

cat("Fine tuning: remove variables with p > 0.05")
mnt.bkw.tune <- lm(newtime ~ group + Tracheotomy + Lung.contusion, 
                   data = dat.t[dat.t$Death != "1",])

summary(mnt.bkw.tune)
lrtest(mnt.bkw, mnt.bkw.tune)
confint(mnt.bkw.tune)
cat("Residual plots of the reduced model")
par(mfrow=c(2,2))
plot(mnt.bkw.tune)

par(mfrow=c(1,1))


g3 <- ggplot(dat.nodeath, aes(x=Tracheotomy, y=newtime, label = pairs)) +
  geom_boxplot() + geom_point(pch=21, size = sz2, aes(fill=group)) + geom_text(nudge_x = xnudge2, size = sz1) +
  labs(x = "Tracheotomy no/yes", y = "days on respirator") 
  
g4 <- ggplot(dat.nodeath, aes(x=Lung.contusion, y=newtime, label = pairs)) +
  geom_boxplot() + geom_point(pch=21, size = sz2, aes(fill=group)) + geom_text(nudge_x = xnudge2, size = sz1) +
  labs(x = "Pulmonary contusion no/yes", y = "days on respirator")

# total time for both groups
ggarrange(g1,g2, nrow = 1, ncol = 2)

# total time for NoSx, newtime for Sx
ggarrange(g3,g4, nrow = 1, ncol = 2)

```



## Linear regression analysis of groups individually

Approach: For each group (Sx or noSx) try to find variables that have nn influence on the time on the respirator in order to identify possible markers for a good or a poor course of disease.    

NOTE: due to the small number of observations these results have to be interpreted with caution.

### Data from noSx group only, y = total time

```{r nobs_nSx, eval=T, echo=F}
# the number of individuals in nSx group
nobs.nSx <- nrow(d.nSx)
# the number of individuals in nSx group without deceased individuals
d.nSx.ndec <- d.nSx[d.nSx$Death != "1",]
nobs.nSx.ndec <- nrow(d.nSx.ndec)

```


The noSx group consists of only `r nobs.nSx.ndec` individuals when deceased 
patients are excluded. This limits the number of parameters that we can 
estimate significantly.


```{r no_sx, eval=T, echo=F}
# we limit the number of parameters 
mnS.full <- lm(Respirator.time ~ Tracheotomy + frac.Sternum + Lung.contusion + 
                 Pneumonia, data = d.nSx.ndec)
summary(mnS.full)

mnS.bkw <- step(mnS.full)
summary(mnS.bkw)

cat("Fine tuning: remove variables with p > 0.05")
mnS.bkw.tune <- lm(Respirator.time ~ Tracheotomy, data = d.nSx.ndec)
summary(mnS.bkw.tune)
lrtest(mnS.full, mnS.bkw.tune)
confint(mnS.bkw.tune)

par(mfrow = c(2,2))
cat("Residual plots of the reduced model")
plot(mnS.bkw.tune)

par(mfrow = c(1,1))

p.trach <- ggplot(d.nSx.ndec, aes(x=Tracheotomy, y=Respirator.time, label = pairs)) + 
  geom_boxplot() + geom_point(size = sz2) + geom_text(nudge_x = xnudge2, size = sz1) +
  labs(x = "Tracheotomy no/yes", y = "days on respirator")

# p.pneu <- ggplot(d.nSx.ndec, aes(x=Pneumonia, y=Respirator.time, label = pairs)) + 
#   geom_boxplot() + geom_point(size = sz2) + geom_text(nudge_x = xnudge2, size = sz1) +
#   labs(x = "Pneumonia no/yes", y = "days on respirator")


ggarrange(p.trach, nrow = 1, ncol = 1)



```



### Data from Sx group only, y = time after surgery (newtime)

```{r nobs_Sx, eval=T, echo=F}
# the number of individuals in nSx group
nobs.Sx <- nrow(d.Sx)
# the number of individuals in nSx group without deceased individuals
d.Sx.ndec <- d.Sx[d.Sx$Death != "1",]
nobs.Sx.ndec <- nrow(d.Sx.ndec)

```


The Sx group consists of `r nobs.Sx.ndec` individuals when deceased patients 
are excluded.


```{r sx_all, eval=T, echo=F}

mS.full <- lm(newtime ~ age + GCS + female + Tracheotomy + frac.Sternum + 
                Lung.contusion + Pneumonia + Neurology, data = d.Sx.ndec)
summary(mS.full)

par(mfrow = c(2,2))
# backward selection
mS.bkw = step(mS.full, direction = 'backward')
summary(mS.bkw)

lrtest(mS.full, mS.bkw)

confint(mS.bkw)

cat("Residual plots of the reduced model")
par(mfrow = c(2,2))
plot(mS.bkw)


d.Sx.ndec$otl <- ifelse(d.Sx.ndec$pairs == "I", 0,1)
d.Sx.ndec$otl <- as.factor(d.Sx.ndec$otl)

xnudge1 = 1.4
xnudge2 = 0.1
sz1 = 3
sz2 = 1.2

p.age <- ggplot(d.Sx.ndec, aes(x=Resp.Time.after.OR, y=age, label = pairs, colour = otl)) + 
  geom_text(nudge_x = xnudge1, size = sz1) + geom_point(size = sz2) + 
  theme(legend.position = "none")

p.fem <- ggplot(d.Sx.ndec, aes(x=female, y=Resp.Time.after.OR)) + 
  geom_boxplot() + geom_point(pch=21, size = sz2, aes(fill=otl)) + 
  geom_text(aes(label = pairs, colour = otl), nudge_x = xnudge2, size = sz1) + 
  theme(legend.position = "none")

p.stern <- ggplot(d.Sx.ndec, aes(x=frac.Sternum, y=Resp.Time.after.OR, label = pairs)) + 
  geom_boxplot() + geom_point(pch=21, size = sz2, aes(fill=otl)) + 
  geom_text(aes(label = pairs, colour = otl), nudge_x = xnudge2, size = sz1) + 
  theme(legend.position = "none")

p.lung <- ggplot(d.Sx.ndec, aes(x=Lung.contusion, y=Resp.Time.after.OR, label = pairs)) + 
  geom_boxplot() + geom_point(pch=21, size = sz2, aes(fill=otl)) + 
  geom_text(aes(label = pairs, colour = otl), nudge_x = xnudge2, size = sz1) +  
  theme(legend.position = "none")

p.pneu <- ggplot(d.Sx.ndec, aes(x=Pneumonia, y=Resp.Time.after.OR, label = pairs)) + 
  geom_boxplot() + geom_point(pch=21, size = sz2, aes(fill=otl)) + 
  geom_text(aes(label = pairs, colour = otl), nudge_x = xnudge2, size = sz1) +  
  theme(legend.position = "none")

p.trach <- ggplot(d.Sx.ndec, aes(x=Tracheotomy, y=Resp.Time.after.OR, label = pairs)) + 
  geom_boxplot() + geom_point(pch=21, size = sz2, aes(fill=otl)) + 
  geom_text(aes(label = pairs, colour = otl), nudge_x = xnudge2, size = sz1) +  
  theme(legend.position = "none")

p.neuro <- ggplot(d.Sx.ndec, aes(x=Neurology, y=Resp.Time.after.OR, label = pairs)) + 
  geom_boxplot() + geom_point(pch=21, size = sz2, aes(fill=otl)) + 
  geom_text(aes(label = pairs, colour = otl), nudge_x = xnudge2, size = sz1) +  
  theme(legend.position = "none")

p.gcs <- ggplot(d.Sx.ndec, aes(x=Resp.Time.after.OR, y=GCS, label = pairs, colour = otl)) + 
  geom_text(nudge_x = xnudge1, size = sz1) + geom_point(size = sz2) + 
  theme(legend.position = "none")

ggarrange(p.age, p.fem, p.stern, p.lung, p.pneu, p.trach, p.neuro, p.gcs, nrow = 3, ncol = 3)

```

### Data from Sx group only, individual I excluded, y = time after surgery (newtime)
Patient I behaves differently than the rest of the Sx group: Patient I suffered from a lung contusion, spent the longest period of the entire group at the respirator (24 days) after surgery and received a tracheotomy, but had a GCS of 15 at admission to the hospital and no complications. Medical records revealed that this patient underwent a combined surgery (Rippenstabilisation und 3-fach Bypassoperation) and suffered from a ‘post-operativem Delir’ which was most probably induced by the long and complex surgery. The following analysis excludes patient I.


```{r Sx_removeOurlier, eval=T, echo=F}

d.Sx.noI <- d.Sx.ndec[-c(10),]

mS.nI.full <- lm(newtime ~ age + GCS + female + Tracheotomy + frac.Sternum + 
                   Lung.contusion + Pneumonia + Neurology, data = d.Sx.noI)
summary(mS.nI.full)

mS.nI.bkw = step(mS.nI.full, direction = 'backward')
summary(mS.nI.bkw)

cat("Fine tuning: remove variables with p > 0.05")
mS.nI.bkw.tune <- lm(formula = newtime ~ GCS + Tracheotomy + Pneumonia, data = d.Sx.noI)
summary(mS.nI.bkw.tune)
lrtest(mS.nI.full, mS.nI.bkw.tune)

confint(mS.nI.bkw.tune)

cat("Residual plots of the reduced model")
par(mfrow = c(2,2))
plot(mS.nI.bkw.tune)


p.age <- ggplot(d.Sx.noI, aes(x=Resp.Time.after.OR, y=age, label = pairs)) + 
  geom_text(nudge_x = xnudge1, size = sz1) + geom_point(size = sz2) + 
  theme(legend.position = "none")

p.fem <- ggplot(d.Sx.noI, aes(x=female, y=Resp.Time.after.OR)) + 
  geom_boxplot() + geom_point(pch=21, size = sz2) + 
  geom_text(aes(label = pairs), nudge_x = xnudge2, size = sz1) + 
  theme(legend.position = "none")

p.stern <- ggplot(d.Sx.noI, aes(x=frac.Sternum, y=Resp.Time.after.OR, label = pairs)) + 
  geom_boxplot() + geom_point(pch=21, size = sz2) + 
  geom_text(aes(label = pairs), nudge_x = xnudge2, size = sz1) + 
  theme(legend.position = "none")

p.lung <- ggplot(d.Sx.noI, aes(x=Lung.contusion, y=Resp.Time.after.OR, label = pairs)) + 
  geom_boxplot() + geom_point(pch=21, size = sz2) + 
  geom_text(aes(label = pairs), nudge_x = xnudge2, size = sz1) +  
  theme(legend.position = "none")

p.pneu <- ggplot(d.Sx.noI, aes(x=Pneumonia, y=Resp.Time.after.OR, label = pairs)) + 
  geom_boxplot() + geom_point(pch=21, size = sz2) + 
  geom_text(aes(label = pairs), nudge_x = xnudge2, size = sz1) +  
  theme(legend.position = "none")

p.trach <- ggplot(d.Sx.noI, aes(x=Tracheotomy, y=Resp.Time.after.OR)) + 
  geom_boxplot() + geom_point(pch=21, size = sz2) + 
  geom_text(aes(label = pairs), nudge_x = xnudge2, size = sz1) +  
  theme(legend.position = "none")

p.neuro <- ggplot(d.Sx.noI, aes(x=Neurology, y=Resp.Time.after.OR, label = pairs)) + 
  geom_boxplot() + geom_point(pch=21, size = sz2) + 
  geom_text(aes(label = pairs), nudge_x = xnudge2, size = sz1) +  
  theme(legend.position = "none")

p.gcs <- ggplot(d.Sx.noI, aes(x=Resp.Time.after.OR, y=GCS, label = pairs)) + 
  geom_text(nudge_x = xnudge1, size = sz1) + geom_point(size = sz2) + 
  theme(legend.position = "none")

ggarrange(p.age, p.fem, p.stern, p.lung, p.pneu, p.trach, p.neuro, p.gcs, nrow = 3, ncol = 3)

```

```{r sessioninfo, eval=T , echo=F}

sessionInfo()

```

